 

### 2023-05-24 markdown/methods.Rmd
normalization validation with mutual information included
to do:
include in thesis

#### Batch correction 
Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types. Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter. To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch. In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect. By the method of moments (= the mean and the variance) the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches. 
Furthermore the ComBat function allows the user to pass a desired effect (the actual biological variance) to be crystallized out from the non-biological variance. This is done by a constructing a design matrix which is also passed to the bayesian framework and can be interpreted as the apriori information about our dataset.

@@

##### Differential expression

To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}. This package employs a linear model approach to determine the fold expression between sample groups. Before initiating the analysis, two matrices need to be computed. The design matrix identifies samples based on their sample type and defines the experimental design. An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface. The expression matrix contains intensities for each identified protein and is obtained at the end of the pipeline. By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.

@@

Next the statistical module of the program performs an empirical Bayes model utilizing the eBayes \citep{Smyth2004} function of the Limma package.

>>>>

#### Batch correction 
Since mass spectrometry experiments are highly influenced by several factors, such as sample processing, the technician, manufacturer and production of the device, the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types when performing dimensionality reduction. A common misconception in biology is, that batches need to be corrected before differential expression analysis, which is the final aim of the developed pipeline. It appears convenient to remove the batch effect before performing any statistical analysis, which is in turn prone to errors especially if the sample groups are not split equally between batches. In this case group differences influence the batch effect reducing statistical power after the correction., the both effects are so called interdependent. The opposite could happen if the experimental design is heavily unbalanced leading to group differences induced by batch correction. 
Finally it is recommended to include the batch factor should in the linear model as a co factor and statistically quantify it. \citep{Nygaard2016}
However it could be of interest to observe biological relevance of the factor of interest in the dimensionality reduction visualization, so a batch correction option was included to meet this need.
To address the non-biological experimental variations, also known as technical factors, the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. 
The expression of a gene $g$ for the batch $m_i$ and sample $j$ can be defined as  $Y_{ijg} = a_{g} + X\beta_{g} + \gamma_{ig} + \delta_{ig}\epsilon_{ijg}$, where $\gamma$ and $\delta\epsilon$ represent the systematic error caused by the batch. 
The Bayesian framework is employed to remove this systematic error by shrinking the effect and pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. Before applying batch correction, the data is standardized using Z-transformation. By the method of moments (= the mean and the variance) the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature driving statistical moments at a comparable level between batches.
The ComBat function allows the user to specify a desired effect, which represents the actual biological variance to be preserved while removing non-biological variance. This is achieved by constructing a design matrix, which is provided as prior information about the dataset within the Bayesian framework. However, in cases of poor experimental design, the factors contributing to unwanted and desired variation may be confounded. In the worst case, if only one sample type per run is observed, it becomes challenging for the program to distinguish between batch-induced variance and biological variance. In such cases, the user is advised to adjust the experimental layout for future experiments. 
The detection algorithm for confounding effects is facilitated by QR matrix decomposition. The matrix $A=QR$ is decomposed into an orthogonal matrix $Q$ and an upper triangular matrix $R$. 
By comparing the rank of matrix $A$ with the number of biologically relevant factors, it is possible to determine the presence of confounding effects in the dataset. The rank is defined as the maximum number of linearly independent columns. The same approach is used in the limma package \citep{Phipson2016} to identify non-solvable coefficients in experimental designs.

@@

##### Differential expression 
To test for differential expression, the analysis utilizes the "limma" package from R Bioconductor \citep{Phipson2016}. This package employs a linear model approach to determine the fold expression between sample groups. Before initiating the analysis, two matrices need to be computed. The design matrix identifies samples based on their sample type and defines the experimental design. An algorithm within the program's backend logic is utilized to automatically generate this matrix according to the user's selection in the interface. The expression matrix contains intensities for each identified protein and is obtained at the end of the pipeline. By calling a function with the design and expression matrices as arguments, the contrast matrix is calculated, enabling the user to decide which comparisons to consider and also perform tests on multiple factors.

@@

Next the statistical module of the program performs an empirical Bayes model utilizing the eBayes \citep{Smyth2004} function of the Limma package. The eBayes function utilizes a moderated t-statistics following t-distributed values. The advantage of this method over the posterior odds is the reduction of hyperparameters to estimate used for the modelling process, also called shrinkage. 
In this context, the hyperparameters to estimate are the coefficient $\beta_{gj}$ for gene $g$ and sample $i$, as well as the variance $\sigma^2_g$ across genes $g$. In the Bayesian approach, probabilities are updated after obtaining new data, referred to as conditional probabilities. For estimation, it is assumed that $\frac{1}{\sigma^2_g} \sim \frac{1}{d_{0}s^2_{0}} \chi^2_{d_{0}}$, representing the prior information of the model.
The probability that $\beta_{gj} \neq 0$ is denoted as $p_j$, where $p_j$ represents the expected proportion of truly differentially expressed genes. The expected distribution of log-fold changes follows the distribution $\beta_{gj} | \sigma^2_g , \beta_{gj} \neq 0 \sim N(0, v_{0j}\sigma^2_g)$.
The posterior mean of $\sigma^2_g$ given $s^2_g$ is calculated as $s^{-2}_g = E(\sigma^2_g | s^2_g) = \frac{d{0}s^2_{0} + d_{g}s^2_{g}}{d_{0} + d_{g}}$. This is where the shrinkage occurs, as the posteriors affect the prior values based on their respective sizes and degrees of freedom.
The moderated t-statistic is defined as $\tilde{t}_{gj} = \frac{\hat{\beta}_{gj}}{\hat{s}_{g}\sqrt{v_{gj}}}$. Once the values of $\tilde{t}$ and $s^{2}$ are calculated, the posterior odds are computed, providing the odds that a particular gene is differentially expressed. The odds for gene $g$ being differentially expressed are denoted as $O_{gj} = \frac{p(\beta_{gj} \neq 0 | \tilde{t}_{gj}, s^2{g})}{p(\beta_{gj} = 0 | \tilde{t}_{gj}, s^2{g})}$, where the numerator represents the probability that the gene is differentially expressed, and the denominator represents the probability that the gene is not differentially expressed \citep{Smyth2004}.

When observing multiple batches at once, as already explained in the chapter batch correction this effect needs to be taken into consideration when performing differential expression analysis. This is done by the multifactor option in the user interface by selecting the batch as a co factor and inclusion in the design matrix.
